Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-01-31 CAV 2025 TODO/Paper_noProof.tex
intro

        To handle complex instances, we revisit a divide-and-conquer approach to break down the complexity: instead of few complex BaB calls, we rely on many small {\em partial} MILP calls. The crucial step is to select very few but very important ReLUs to treat using (costly) binary variables. The previous attempts were suboptimal in that respect. In this paper, we propose a novel {\em solution-aware} ReLU scoring (SAS), as well as adapt the BaB-SR and BaB-FSB branching functions as {\em global} ReLU scoring (GS) functions. We compare them theoretically as well as experimentally. Surprisingly perhaps, the most accurate solution (SAS) for {\em selecting} ReLUs to treat as binary variables is different from the most efficient solution (GS) to {\em rank} within this selection. Compared with previous attempts, SAS reduces the number of integer variables by around 4 times while maintaining the same level of accuracy. Implemented in {\em Hybrid MILP}, calling first $\alpha,\beta$-Crown with a short time-out to solve easier instances, and then partial MILP, produces a very accurate yet efficient verifier, reducing by up to $40\%$ the number of undecided instances to low levels ($8-15\%$), while keeping a reasonable runtime ($46s-417s$ on average per instance), even for fairly large CIFAR CNNs with 20 000 nodes.

>>>>

        To handle complex instances, we revisit a divide-and-conquer approach to break down the complexity: instead of few complex BaB calls, we rely on many small {\em partial} MILP calls. The crucial step is to select very few but very important ReLUs to treat using (costly) binary variables. The previous attempts were suboptimal in that respect. In this paper, we propose a novel {\em solution-aware} ReLU scoring (SAS), as well as adapt the BaB-SR and BaB-FSB branching functions as {\em global} ReLU scoring (GS) functions. We compare them theoretically as well as experimentally. Surprisingly perhaps, the most accurate solution (SAS) for {\em selecting} ReLUs to treat as binary variables is different from the most efficient solution (GS) to {\em branch} within this selection. Compared with previous attempts, SAS reduces the number of binary variables by around 6 times while maintaining the same level of accuracy. Implemented in {\em Hybrid MILP}, calling first $\alpha,\beta$-Crown with a short time-out to solve easier instances, and then partial MILP, produces a very accurate yet efficient verifier, reducing by up to $40\%$ the number of undecided instances to low levels ($8-15\%$), while keeping a reasonable runtime ($46s-417s$ on average per instance), even for fairly large CIFAR CNNs with 20 000 nodes.

### 2025-01-31 CAV 2025 TODO/introduction.tex
intro

	\item On the first question, we adapted from BaB-SR \cite{BaB} and FSB \cite{FSB}, which choose branching nodes for BaB, global scoring (GS) functions to choose ReLU for pMILP. 
	These revealed to be much more accurate than the simple heuristic in 
	\cite{DivideAndSlide}. We uncover however one case (out of 4) where the exact function that GS tries to approximate depends heavily upon the mode of the ReLU function, giving very different result for each mode. This implies that these strategies may not be that accurate, because they do not know the mode to choose from. We then designed a novel solution-aware scoring (SAS), which uses the solution of a unique LP call, that provides the mode to consider. Theoretically, we show that SAS is always an over-approximation of the real improvement (Proposition \ref{prop2}), which implies that a small SAS value implies that the ReLU is unnecessary. Experimentally, we further show that SAS is very close to the actual improvement, closer than GS, and that overall, the accuracy from SAS is significantly better than GS.  Compared with the heuristic in \cite{DivideAndSlide},  SAS is much more efficient ($\approx 6$ times less integer variables for same accuracy (Plot \ref{fig_table3})).We however explain why focusing on a solution may not be appropriate to choose branching order within this selection, and that GS would be better suited, which we verify experimentally.

@@

	\item For the second question, we propose a new verifier, called {\em Hybrid MILP}, invoking first 	$\alpha,\beta$-Crown with short time-out to settle the easy instances. On those ({\em hard}) instances which are neither certified nor falsified, we call pMILP with few neurons encoded as integer variables. Experimental evaluation reveals that Hybrid MILP achieves a beneficial balance between accuracy and completeness compared to prevailing methods. It reduces the proportion of undecided inputs from $20-58\%$ ($\alpha,\beta$-Crown with 2000s TO) to $8-15\%$, while taking a reasonable average time per instance ($46-420$s), Table \ref{table_hybrid}. It scales to fairly large networks such as CIFAR-10 CNN-B-Adv \cite{SDPFI}, with more than 20 000 neurons.
%We verify experimentally that the algorithm offers interesting trade-offs, by testing on local robustness for DNNs trained "naturally" (and thus difficult to verify).
%KSM: I think this is a distraction in intro, so I suggest moving to later part
% Overall, the worst case complexity of algorithm \ref{algo1} is lower than $O(N 2^K LP(N))$, where $N$ is the number of nodes of the DNN, $K$ the number of ReLU nodes selected as binary variable, and $LP(N)$ is the (polynomial time) complexity of solving a linear program representing a DNN with $N$ nodes. This complexity is an upper bound, as e.g. Gurobi is fairly efficient and never need to consider all of the $2^K$ ReLU configurations to compute the bounds. Keeping $K$ reasonably low thus provides an efficient algorithm. 
%By design, it will never run into a complexity wall (unlike the full MILP encoding), although it can take a while on large networks because of the linear factor $N$ in the number of nodes.
\end{enumerate}

>>>>

	\item On the first question, we adapted from BaB-SR \cite{BaB} and FSB \cite{FSB}, which choose branching nodes for BaB, global scoring (GS) functions to choose ReLU for pMILP. 
	These revealed to be much more accurate than the simple heuristic in 
	\cite{DivideAndSlide}. We uncover however one case (out of 4) where the exact function that GS tries to approximate depends heavily upon the mode of the ReLU function, giving very different result for each mode. This implies that these strategies may not be that accurate, because they do not know the mode to choose from. We then designed a novel solution-aware scoring (SAS), which uses the solution of a unique LP call, that provides the mode to consider. Theoretically, we show that SAS is always an over-approximation of the real improvement (Proposition \ref{prop2}), which implies that a small SAS value implies that the ReLU is unnecessary. Experimentally, we further show that SAS is very close to the actual improvement, closer than GS, and that overall, the accuracy from SAS is significantly better than GS.  Compared with the heuristic in \cite{DivideAndSlide},  SAS is much more efficient ($\approx 6$ times less binary variables for same accuracy (Plot \ref{fig_table3})). We however explain why focusing on a solution may not be appropriate to choose branching order within this selection, and that GS would be better suited, which we verify experimentally.

@@

	\item For the second question, we propose a new verifier, called {\em Hybrid MILP}, invoking first 	$\alpha,\beta$-Crown with short time-out to settle the easy instances. On those ({\em hard}) instances which are neither certified nor falsified, we call pMILP with few neurons encoded as binary variables. Experimental evaluation reveals that Hybrid MILP achieves a beneficial balance between accuracy and completeness compared to prevailing methods. It reduces the proportion of undecided inputs from $20-58\%$ ($\alpha,\beta$-Crown with 2000s TO) to $8-15\%$, while taking a reasonable average time per instance ($46-420$s), Table \ref{table_hybrid}. It scales to fairly large networks such as CIFAR-10 CNN-B-Adv \cite{SDPFI}, with more than 20 000 neurons.
%We verify experimentally that the algorithm offers interesting trade-offs, by testing on local robustness for DNNs trained "naturally" (and thus difficult to verify).
%KSM: I think this is a distraction in intro, so I suggest moving to later part
% Overall, the worst case complexity of algorithm \ref{algo1} is lower than $O(N 2^K LP(N))$, where $N$ is the number of nodes of the DNN, $K$ the number of ReLU nodes selected as binary variable, and $LP(N)$ is the (polynomial time) complexity of solving a linear program representing a DNN with $N$ nodes. This complexity is an upper bound, as e.g. Gurobi is fairly efficient and never need to consider all of the $2^K$ ReLU configurations to compute the bounds. Keeping $K$ reasonably low thus provides an efficient algorithm. 
%By design, it will never run into a complexity wall (unlike the full MILP encoding), although it can take a while on large networks because of the linear factor $N$ in the number of nodes.
\end{enumerate}
